---
title: "FinalTeam2 HypoTesting"
author: "Final Team: Jocelyn Yang, Yiming Xue, Janel Wu, Yichen Wang, Xinyang Wang"
date: "2024-12-01"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
df <- read.csv("final_data_CHILD.csv")
```

# Hypothesis Testing 
3 Testable Hypotheses along with models
For child-involved crimes:
## Hypothesis 1:
If long & short time delay in Data Occurred & Date Reported results in significantly different Status (AA, IC, AO). This can be categorized as arrested or not to fit logistic regression
Test Method: Logistic regression

```{r}
library(lubridate)
library(randomForest)
# convert time
df$DATE.OCC <- as.POSIXct(df$DATE.OCC, format = "%m/%d/%Y %I:%M:%S %p")
df$Date.Rptd <- as.POSIXct(df$Date.Rptd, format = "%m/%d/%Y %I:%M:%S %p")

# calculate time delay
df$delay <- as.numeric(difftime(df$Date.Rptd, df$DATE.OCC, units = "days"))
df <- df[df$delay >= 0, ]

# categorize into 'short' and 'long' 
median_delay <- median(df$delay, na.rm = TRUE)
df$delay_cat <- ifelse(df$delay <= median_delay, "short", "long")

# recode 'Status.Desc' to 'Arrested' or 'Not Arrested'
df$Arrested <- ifelse(df$Status.Desc %in% c("Adult Arrest", "Juvenile Arrest"), 1, 0)
df$Arrested <- factor(df$Arrested, levels = c(0, 1), labels = c("Not Arrested", "Arrested"))
df_hyp1 <- df[!is.na(df$Arrested) & !is.na(df$delay_cat), ]

# fit logistic regression model
model <- glm(Arrested ~ delay_cat, data = df_hyp1, family = binomial)
summary(model)

exp(cbind(Odds_Ratio = coef(model), confint(model)))
```

Results:
Null Hypothesis (H_0): There is no significant difference in arrest status (arrested or not) between crimes with long and short delays between occurrence and reporting.

Alternative Hypothesis (H_1): There is a significant difference in arrest status between crimes with long and short delays between occurrence and reporting.

Conclusion:
Because the p-value for the delay category is highly significant (p < 0.001), we reject the null hypothesis (H_0) and accept the alternative hypothesis (H_1), suggesting that crimes reported sooner (short delay) are significantly more likely to result in an arrest compared to those with a longer delay.

The odds of an arrest are approximately 48% higher for crimes with a short reporting delay compared to those with a long delay. Prompt reporting of child-involved crimes enhances the likelihood of apprehending the suspect, highlighting the importance of encouraging immediate reporting to improve justice outcomes.

## Hypothesis 2:
If different months are associated with different types of weapon used, to investigate if certain months present more violent crimes.
Test Method: Chi-square.

```{r}
df_hyp2 <- df[!is.na(df$Weapon.Desc), ]
df_hyp2$Month <- as.factor(format(as.Date(df_hyp2$DATE.OCC,
                                          format = "%m/%d/%Y %I:%M:%S %p"), "%b"))

# group months into seasons
df_hyp2$Season <- factor(ifelse(df_hyp2$Month %in% c("Dec", "Jan", "Feb"), "Winter",
                         ifelse(df_hyp2$Month %in% c("Mar", "Apr", "May"), "Spring",
                         ifelse(df_hyp2$Month %in% c("Jun", "Jul", "Aug"), "Summer", "Fall"))),
                         levels = c("Spring", "Summer", "Fall", "Winter"))

weapon_counts <- table(df_hyp2$Weapon.Desc)
rare_weapons <- names(weapon_counts[weapon_counts < 50])  # Adjust threshold as needed
df_hyp2$Weapon.Desc <- as.character(df_hyp2$Weapon.Desc)
df_hyp2$Weapon.Desc[df_hyp2$Weapon.Desc %in% rare_weapons] <- "Other"
df_hyp2$Weapon.Desc <- factor(df_hyp2$Weapon.Desc)

table_season_weapon <- table(df_hyp2$Season, df_hyp2$Weapon.Desc)

chisq_result <- chisq.test(table_season_weapon)

chisq_result
table_season_weapon
```
Results: 
Null Hypothesis (H_0): There is no association between the season in which a crime occurs and the type of weapon used in child-involved crimes.
Alternative Hypothesis (H_1): There is an association between the season and the type of weapon used in child-involved crimes.

Since the p-value is less than 0.05, we reject the null hypothesis (H_0) and accept the alternative hypothesis (H_1), indicating a significant association between seasons and weapon types used.

Conclusion:
There exists a statistically significant association between the season and the type of weapon used in crimes involving victims under 12 years old. The type of weapon used in child-involved crimes varies by season, indicating that certain weapons are more prevalent during specific times of the year.

We observe that Winter shows higher instances of physical assaults, possibly due to more indoor interactions during colder months.


```{r}
# Identify cells with residuals > 2 or < -2
residuals <- chisq_result$stdres

significant_cells <- which(abs(residuals) > 2, arr.ind = TRUE)
significant_cells
```
```{r}
row_names <- rownames(residuals)
col_names <- colnames(residuals)

for (i in 1:nrow(significant_cells)) {
  row_idx <- significant_cells[i, "row"]
  col_idx <- significant_cells[i, "col"]
  
  season <- row_names[row_idx]
  weapon <- col_names[col_idx]
  res_value <- residuals[row_idx, col_idx]
  
  cat(sprintf("Season: %s, Weapon: %s, Residual: %.2f\n", 
              season, weapon, res_value))
}

```
```{r}
library(ggplot2)
library(viridis) 

plot <- ggplot(df_hyp2, aes(x = Season, fill = Weapon.Desc)) +
  geom_bar(position = "dodge") +
  labs(
    title = "Weapon Usage Across Seasons",
    x = "Season",
    y = "Number of Incidents"
  ) +
  theme_minimal() +
  scale_fill_viridis_d(option = "mako") # Use mako color palette

ggsave("weapon_types_by_season.png", plot = plot, width = 8, height = 6, dpi = 300)

print(plot)


```


```{r}
# sex comparison
library(lubridate)
library(ggplot2)
library(reshape2)
library(dplyr)

df_hyp4 <- df
valid_vict_sex <- !is.na(df_hyp4$Vict.Sex) & df_hyp4$Vict.Sex != ""
df_hyp4 <- df_hyp4[valid_vict_sex, ]
df_hyp4 <- df_hyp4[df_hyp4$Vict.Sex != "H",]

df_hyp4$Vict.Sex <- factor(df_hyp4$Vict.Sex)


df_hyp4$Vict.Sex <- droplevels(df_hyp4$Vict.Sex)

```

```{r}
df_hyp4$Weapon.Desc <- as.character(df_hyp4$Weapon.Desc)  
df_hyp4$Weapon.Desc[is.na(df_hyp4$Weapon.Desc) | df_hyp4$Weapon.Desc == ""] <- "Missing"

# Combine less frequent weapon types into 'Other' (threshold < 50 occurrences)
weapon_counts <- table(df_hyp4$Weapon.Desc)
rare_weapons <- names(weapon_counts[weapon_counts < 50])  # Identify rare weapons
df_hyp4$Weapon.Desc[df_hyp4$Weapon.Desc %in% rare_weapons] <- "Other"  

df_hyp4$Weapon.Desc <- factor(df_hyp4$Weapon.Desc)

df_hyp4$DATE.OCC <- as.POSIXct(df_hyp4$DATE.OCC, format = "%m/%d/%Y %I:%M:%S %p")

df_hyp4$Month <- as.factor(format(df_hyp4$DATE.OCC, "%b"))

df_hyp4$Vict.Sex <- as.character(df_hyp4$Vict.Sex)

valid_vict_sex <- !is.na(df_hyp4$Vict.Sex) & df_hyp4$Vict.Sex != ""

df_hyp4 <- df_hyp4[valid_vict_sex, ]

df_hyp4$Vict.Sex <- factor(df_hyp4$Vict.Sex, levels = c("F", "M", "X"))

table_victim_weapon <- table(df_hyp4$Vict.Sex, df_hyp4$Weapon.Desc)

# performing the Chi-Squared Test

chisq_result_victim_weapon <- chisq.test(table_victim_weapon)
print(chisq_result_victim_weapon)
print(table_victim_weapon)
```

```{r}
# Identify cells with residuals > 2 or < -2
residuals <- chisq_result_victim_weapon$stdres

significant_cells <- which(abs(residuals) > 2, arr.ind = TRUE)
significant_cells

row_names <- rownames(residuals)
col_names <- colnames(residuals)

# Loop through significant cells and print details
for (i in 1:nrow(significant_cells)) {
  row_idx <- significant_cells[i, "row"]
  col_idx <- significant_cells[i, "col"]
  
  season <- row_names[row_idx]
  weapon <- col_names[col_idx]
  res_value <- residuals[row_idx, col_idx]
  
  cat(sprintf("Sex: %s, Weapon: %s, Residual: %.2f\n", 
              season, weapon, res_value))
}
```


```{r}
library(ggplot2)
library(scales)   # For percent_format()
library(viridis)  

# stacked bar plot
plot <- ggplot(df_hyp4, aes(x = Vict.Sex, fill = Weapon.Desc)) +
  geom_bar(position = "fill") +  # Stacked bars scaled to 1 (proportions)
  labs(
    title = "Proportion of Weapon Types by Victim Sex",
    x = "Victim Sex",
    y = "Proportion of Incidents",
    fill = "Weapon Description"
  ) +
  scale_y_continuous(labels = percent_format()) +  # Convert y-axis to percentages
  theme_minimal() +  # Use a clean, minimal theme
  theme(
    plot.title = element_text(hjust = 0.5)  # Center the title
  ) +
  scale_fill_viridis_d(option = "mako")  # Use the mako color palette

print(plot)

ggsave(
  filename = "proportion_weapon_types_by_victim_sex.png", 
  plot = plot, 
  width = 8, 
  height = 6, 
  dpi = 300
)
```



# Hypothesis 3
The type of crime (Crm.Cd.Desc) can be predicted by location type (Premis.Desc), time of occurrence (Time Occ), and victim demographics (Vict Age, Vict Sex, etc.).
Test Method: Random forest & Variable selection. Encode categorical predictors to numeric representations.

```{r}
# Hypothesis 3: 
library(randomForest)

df_hyp3 <- df[, c("Crm.Cd.Desc", "Premis.Desc", "TIME.OCC", 
                  "Vict.Age", "Vict.Sex", "Vict.Descent")]
df_hyp3 <- na.omit(df_hyp3)

# Encode categorical variables as factors
df_hyp3$Crm.Cd.Desc <- as.factor(df_hyp3$Crm.Cd.Desc)
df_hyp3$Premis.Desc <- as.factor(df_hyp3$Premis.Desc)
df_hyp3$Vict.Sex <- as.factor(df_hyp3$Vict.Sex)
df_hyp3$Vict.Descent <- as.factor(df_hyp3$Vict.Descent)

# Convert 'TIME.OCC' to time categories
df_hyp3$Time.Category <- cut(df_hyp3$TIME.OCC, breaks = c(-1, 600, 1200, 1800, 2400),
                             labels = c("Night", "Morning", "Afternoon", "Evening"))
df_hyp3$Time.Category <- as.factor(df_hyp3$Time.Category)
df_hyp3$TIME.OCC <- NULL 

# Combine similar crime types and group similar crimes into broader categories
df_hyp3$Crime.Category <- as.character(df_hyp3$Crm.Cd.Desc)
assault_crimes <- c("ASSAULT WITH DEADLY WEAPON, AGGRAVATED ASSAULT",
                    "BATTERY - SIMPLE ASSAULT",
                    "CHILD ABUSE (PHYSICAL) - SIMPLE ASSAULT")
neglect_crimes <- c("CHILD NEGLECT (SEE 300 W.I.C.)",
                    "CRM AGNST CHLD (13 OR UNDER) (14-15 & SUSP 10 YRS OLDER)")

df_hyp3$Crime.Category[df_hyp3$Crime.Category %in% assault_crimes] <- "Assault"
df_hyp3$Crime.Category[df_hyp3$Crime.Category %in% neglect_crimes] <- "Neglect/Abuse"

# Remove less frequent crime types
df_hyp3 <- df_hyp3[df_hyp3$Crime.Category %in% c("Assault", "Neglect/Abuse"), ]
df_hyp3$Crime.Category <- as.factor(df_hyp3$Crime.Category)
df_hyp3$Crm.Cd.Desc <- NULL

# Combine less frequent levels in 'Premis.Desc'
premis_counts <- table(df_hyp3$Premis.Desc)
rare_premises <- names(premis_counts[premis_counts < 100])  # Adjust threshold as needed
df_hyp3$Premis.Desc <- as.character(df_hyp3$Premis.Desc)
df_hyp3$Premis.Desc[df_hyp3$Premis.Desc %in% rare_premises] <- "Other"
df_hyp3$Premis.Desc <- as.factor(df_hyp3$Premis.Desc)

# Show num of levels
sapply(df_hyp3, function(x) if(is.factor(x)) nlevels(x) else NA)

# test train split
set.seed(123)
train_indices <- sample(1:nrow(df_hyp3), size = 0.7 * nrow(df_hyp3))
train_data <- df_hyp3[train_indices, ]
test_data <- df_hyp3[-train_indices, ]

# Fit the rf model with adjusted hyperparameters
rf_model <- randomForest(Crime.Category ~ ., data = train_data, mtry = 2, 
                         ntree = 300, importance = TRUE)
predictions <- predict(rf_model, newdata = test_data)

# Confusion matrix
confusion_matrix <- table(Predicted = predictions, Actual = test_data$Crime.Category)
confusion_matrix

# Calculate accuracy
accuracy <- sum(diag(confusion_matrix)) / sum(confusion_matrix)
print(paste("Accuracy:", round(accuracy * 100, 2), "%"))

```
Result: 
Null Hypothesis (H_0): The type of crime (assault vs. neglect/abuse) cannot be predicted by location type, time of occurrence, and victim demographics better than random chance.

Alternative Hypothesis (H_1): The type of crime can be predicted by location type, time of occurrence, and victim demographics better than random chance.

Conclusion:
The model correctly classified crimes as "Assault" or "Neglect/Abuse" 66.75% of the time. Note that this accuracy is significantly higher than the baseline accuracy of 50% expected from random guessing between two classes. Because the model predicts the crime type significantly better than random chance, we reject the null hypothesis (H_0). This indicates that the type of crime can be predicted by location type, time of occurrence, and victim demographics.

The results showed that even within the 1-12 year range, certain ages may be more susceptible to specific crimes. Gender and descent may play a role in susceptibility to crime type.

```{r}
# Install randomForest if it's not already installed
if (!require(randomForest)) {
  install.packages("randomForest")
  library(randomForest)
} else {
  library(randomForest)
}

```


```{r}
# Variable Importance Plot using base randomForest function
varImpPlot(rf_model, 
           main = "Variable Importance",
           type = 2)  # type=2 for Mean Decrease Gini

# Alternatively, using ggplot2 for a more customized plot
library(ggplot2)
importance_df <- as.data.frame(importance(rf_model))
importance_df$Variable <- rownames(importance_df)

ggplot(importance_df, aes(x = reorder(Variable, MeanDecreaseGini), y = MeanDecreaseGini)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  coord_flip() +
  labs(title = "Variable Importance (Mean Decrease Gini)",
       x = "Variables",
       y = "Importance") +
  theme_minimal()

```

```{r}
# Install and load required packages
if (!require("ggplot2")) install.packages("ggplot2")
if (!require("reshape2")) install.packages("reshape2")

library(ggplot2)
library(reshape2)

# Assuming confusion_matrix is already created as in your code
confusion_matrix <- table(Predicted = predictions, Actual = test_data$Crime.Category)

# Convert the confusion matrix to a data frame
confusion_df <- as.data.frame(confusion_matrix)

# Plot heatmap
ggplot(confusion_df, aes(x = Actual, y = Predicted, fill = Freq)) +
  geom_tile(color = "white") +
  scale_fill_gradient(low = "lightblue", high = "steelblue") +
  geom_text(aes(label = Freq), color = "white", size = 4) +
  labs(title = "Confusion Matrix Heatmap",
       x = "Actual",
       y = "Predicted") +
  theme_minimal()

```
```{r}
# Install and load required package
if (!require("pdp")) install.packages("pdp")
library(pdp)

# Example: Partial Dependence for 'Vict.Age'
partial_plot_age <- partial(rf_model, 
                            pred.var = "Vict.Age", 
                            train = train_data)
plotPartial(partial_plot_age, main = "Partial Dependence of Vict.Age")

# Example: Partial Dependence for 'Time.Category'
partial_plot_time <- partial(rf_model, 
                             pred.var = "Time.Category", 
                             train = train_data)
plotPartial(partial_plot_time, main = "Partial Dependence of Time.Category")

```


